Concurrent Programming with CXL

### 2020-05-29 book/paper.tex
Various updates

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

CXL is not an object-oriented language, however, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If \texttt{d} is a
dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len~d} returns the size of
this set.

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has a method \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
method returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

Figure~\ref{fig:csbarebones} shows the code.
The assertion also makes use of the method \texttt{nametag()} that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

However, mutual exclusion by itself easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\begin{verbatim}
while True: pass;;
\end{verbatim}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

Mutual exclusion is an example of a \emph{safety property}, a
property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}: we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if some set of processes
$S$ are trying to enter the critical section, then eventually one process in
$S$ will enter the critical section if any process already in the
intersection eventually leaves, even if no processes other than $S$ ever
try to enter the critical section.
We call this \emph{progress}.

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
Note also that we have dropped the \texttt{@ncs} label: it would serve no
specific purpose.

@@

Figure~\ref{fig:uplock} presents a solution based on a na\"{\i}ve implementation
of a \emph{lock}.
The idea is that the lock is a baton that at most one process can own at
a time.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until lock is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then flags[i] == True.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Suppose that process~1 is at the critical section from $t_2$ until
it clears its flag at time $t_3$.  Note $t_0 < t_1 < t_2 < t_3$.
Process~1 can only reach the critical section between $(t_2, t_3)$
if it finds \texttt{flags[0] == False} during $(t_1, t_3)$.  But
during $(t_0, t_3)$ \texttt{flags[0] == True}.

However, the solution has a different problem called \emph{livelock}:
if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violated \emph{progress}.

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is no where near the critical section
or has given way.
Running the algorithm though CXL shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly satisfies ME1 and ME2.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant and therefore does
not satisfy ME3.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
The figure uses the CXL \texttt{atomic} statement for the first time.
For now, all you need to know is that it turns the embedded statement into
a single atomic step.  The label \texttt{@gate} thus refers exactly to
this statement.  You can run Figure~\ref{fig:petersonproof} to determine
that $I_p(i)$ is indeed an invariant for $i = 0, 1$.

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language through its \texttt{atomic}
statement.
You already saw an example of this in Figure~\ref{fig:petersonproof}.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{verbatim}
\end{code}

@@

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False} and no other process modifies
\texttt{private[$i$]}.

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the variable \texttt{sum}
is used to compute the number of variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
Note that this process must be started after the variables are
initialized---before that the invariant does not hold yet.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

In CXL, atomic statements allow you to \emph{implement} your own
low synchronization primitives like test-and-set.  Atomic statements
are not intended to \emph{replace} locks or other synchonization primitives.
By the way, CXL \texttt{assert} statements execute atomically
as if they are nested within an atomic statement.

@@

Locks are probably the most popular and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:locks} shows how they are implemented, and
Figure~\ref{fig:figinc} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

Locks are useful when accessing a shared data structure.  By preventing
more than one process from accessing the data structure at the same
time, conflicting access are avoided.  However, not all concurrent
accesses conflict, and opportunities for concurrency may be lost,
hurting performance.  One important case is when multiple processes
are simply reading the data structure.  There is no conflict, and
all reads should be allowed to proceed concurrently.  In many
applications, reads are the majority of all accesses.

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The lock is used to protect access to those counters.
The program shows two kinds of processes: \texttt{reader()} and
\texttt{writer()}.  The critical section is spread between two
labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} \ge n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $\exists$ writer at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = 0$,
\end{itemize}

>>>>

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps CXL values, known as \emph{keys}, to CXL values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

CXL is not an object-oriented language, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If \texttt{d} is a
dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len~d} returns the size of
this set.

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has an operator \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
operator returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

Figure~\ref{fig:csbarebones} shows the code.
The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

However, mutual exclusion by itself easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\begin{verbatim}
while True:
    pass;
;
\end{verbatim}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

Mutual exclusion is an example of a \emph{safety property}, a
property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}: we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if some set of processes
$S$ are trying to enter the critical section and any process already in the
intersection eventually leaves, then eventually one process in $S$ will enter
the critical section.
We call this \emph{progress}.

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/mutex.png}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
Note also that we have dropped the \texttt{@ncs} label: it would serve no
specific purpose.
Figure~\ref{fig:mutex} show a high-level state diagram of the code in
Figure~\ref{fig:cs}.

@@

In order to have at most one process in a critical section at a time, you
probably are thinking that one might use a \emph{lock}.
That is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve implementation of a lock.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then flags[i] == True.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \texttt{flags[1]} to \texttt{True} and
then wait until \texttt{flags[0] == False}.  But that does not happen
until time $t_2$.

However, the solution has a different problem called \emph{livelock}:
if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violates \emph{progress}.

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is no where near the critical section
or has given way.
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly satisfies ME1 and ME2.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant and therefore does
not satisfy ME3.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
The label \texttt{@gate} refers to the instruction that sets \texttt{turn} to $1-i$.
You can run Figure~\ref{fig:petersonproof} to determine
that $I_p(i)$ is indeed an invariant for $i = 0, 1$.

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{verbatim}
\end{code}

@@

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the variable \texttt{sum}
is used to compute the number of variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
In CXL, \texttt{assert} statements are executed atomically, so the
evaluation of the assertion is not interleaved with the execution
of other processes.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

In CXL, atomic statements allow you to \emph{implement} your own
low synchronization primitives like test-and-set.  Atomic statements
are not intended to \emph{replace} locks or other synchonization primitives.

@@

Locks are probably the most popular and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:locks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/rdwr.png}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

Locks are useful when accessing a shared data structure.  By preventing
more than one process from accessing the data structure at the same
time, conflicting accesses are avoided.  However, not all concurrent
accesses conflict, and opportunities for concurrency may be lost,
hurting performance.  One important case is when multiple processes
are simply reading the data structure.
In many applications, reads are the majority of all accesses.
Allowing reads to proceed concurrently can significantly improve performance.

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The lock is used to protect access to those counters.
The program shows a process that executes in a loop.
Each time, it decides whether to read or write.
The critical section is spread between two labels:
readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
specification for two processes.

\begin{quote}
\begin{itemize}
\item Draw additional states and steps in Figure~\ref{fig:rdwr} to 
for three processes.
\end{itemize}
\end{quote}

A process that wants to read first waits until there are no writers:
\texttt{nwriters == 0}.  If so, it increments the number of readers.
Similarly, a
process that wants to write waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathtt{nwriters} \ge n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathtt{nreaders} = 0$,
\end{itemize}
